---
title: "Forecasting for Tanga Sample Data"
subtitle: "with complete facility data"
author: "Peter Boshe"
format: 
  html:
    fig-width: 8
    fig-height: 4
    code-fold: true
    toc: true
    toc-depth: 5
    toc-location: left
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```



## Business Problem(s)
- forecast the amc so as to calculate the amount needed to restock to 4 months
- forecast the amount of product to be ordered

Assumptions;

- 95% confidence interval

### Methodology

Things to do;

- we will be forecasting for amc
- use other variables to check for outliers and other analysis
- Forecasting will be done for different levels 
  - by individual group
  - by facility
  - by facility types
  - by different ordering periods(bi-monthly, in 4 month aggregates, annually)
- Build a demo shiny app UI with user interactivity
  
### Data Extraction

We downloaded our data from eLMIS R&R bi-monthly reports for Tanga region for the years
2020 to 2022, the records start from the march-april reporting period.


```{r import}
#| warning: false
#| message: false
#| echo: false
library(tidyverse)
library(fpp2)
library(visdat)
library(gt)

#import the dataset
df_path <- here::here("data/tanga_2020_2022.rds")
df_imported <- read_rds(df_path) |> janitor::clean_names()
#set theme for all the plots
theme_set(theme_bw())
```




### Data Cleaning and Preprocessing

#### Pre-processing the data

During the pre-processing process we;

- selected the columns of interest,
- clean the white spaces and ensure that all the monthly data have uniform spelling
- performed factor ordering to make sure that the months are ordered chronologically


```{r pre-process}
#| echo: false

df <- df_imported |> 
      # change each column to the prefered datatype
      transmute(
        line = as.numeric(line_number),
        facility = as.factor(facility),
        # facility_type = as.factor(facility_type),
        district = as.factor(district),
        zone = as.factor(zone),
        period = as.character(period),
        product = product_name,
        amc = as.numeric(amc),
        physical_count = as.numeric(physical_count),
        mos = as.numeric(mos),
        order_quantity = as.numeric(order_quantity),
        status = as.factor(status),
        group = as.factor(group),
        year = as.factor(year)
      ) |>
      separate(period, into = c("month_range","ext_year"), sep = "\\s(?=[^\\s]+$)") |>
      separate(month_range, into = c("start_month", "end_month"), sep = "-") |>
      mutate(
        start_month = trimws(tolower(start_month)),
        end_month = trimws(tolower(end_month)),
        start_month = case_when(
          start_month == "jan" ~ "january",
          start_month == "feb" ~ "february",
          start_month == "mar" ~ "march",
          start_month == "apr" ~ "april",
          start_month == "may" ~ "may",
          start_month == "jun" ~ "june",
          start_month == "jul" ~ "july",
          start_month == "aug" ~ "august",
          start_month == "sept" ~ "september",
          start_month == "oct" ~ "october",
          start_month == "nov" ~ "november",
          start_month == "dec" ~ "december",
          TRUE ~ start_month
        ),
        end_month = case_when(
          end_month == "jan" ~ "january",
          end_month == "feb" ~ "february",
          end_month == "mar" ~ "march",
          end_month == "apr" ~ "april",
          end_month == "may" ~ "may",
          end_month == "jun" ~ "june",
          end_month == "jul" ~ "july",
          end_month == "aug" ~ "august",
          end_month == "sept" ~ "september",
          end_month == "oct" ~ "october",
          end_month == "nov" ~ "november",
          end_month == "dec" ~ "december",
          TRUE ~ end_month
        )
      ) |>
      unite("month_range", c(start_month, end_month), sep = " - ")

    # introduce factor ordering for proper arrangement of the reporting periods
    #
    ## Reordering df$month_range
    df$month_range <- df$month_range %>%
      fct_relevel(
        "january - february", "february - march", "march - april",
        "april - may", "may - june", "june - july", "july - august",
        "august - september", "september - october", "october - november",
        "november - december", "december - january"
      )
    # rearranging the dataset
    #
    df <- df |>
      arrange(year, facility, month_range)
    print("pre-process successful...")


```






### Data Quality

Due to the nature of the data received, The R&R reports seem to have a lot of inconsistency that may be caused by the human element of data recording leading to having multiple strings indicating the same data points.

<!-- #### Summary of the processed data -->

<!-- ```{r} -->
<!-- df |>  -->
<!--   skimr::skim() -->
<!-- ``` -->

#### Check for missing variables

```{r missing}
# check for missing variables
# df |> 
#   vis_miss(warn_large_data = F)

sapply(df, function(x) sum(is.na(x)))

df |> filter(is.na(order_quantity))


```

#### Check for duplicated rows

```{r duplicates}
# check for duplicates
df |> 
  duplicated() |> 
  sum()


```

- we do not have missing data
- we do not have duplicated data


#### Inconsistency in naming periods

```{r}
df_imported |> 
   separate(period, into = c("month_range","ext_year"), sep = "\\s(?=[^\\s]+$)") |>
  count(month_range)

```

- we have a data quality issue where the time periods do not have a uniform conventional naming structure
- before cleaning we have 14 period ranges due to data recording

#### Inconsistency in naming products and product codes

```{r}
df_imported |>
  filter(str_detect(product_name, "Tenofovir")) |>
  distinct(product_name, product_code_name) |> 
  view()
```

- there is a possibility that the product codes might be changed
- there is a possibility of duplicates by renaming the same products
- there is no consistency in product coding and naming (ie. same product name might have different product code like 10011380AB, 10010022AB both named Tenofovir Disoproxil Fumarate +Lamivudine+Doltegravir (TLD) Tablet (300/300/50) mg
)

#### Checking completeness of data

```{r}
length(unique(df$facility))
```


- for data quality checks we need to confirm that the number of facilities match the reality 
- with the new data we have 347 facilities




### Exploratory data analysis (EDA)

#### Distribution of amc and checking for outliers

```{r eda}

# check for outliers

df |> 
  ggplot(aes(amc)) +
  geom_boxplot()

df |> 
  ggplot(aes(log(amc))) +
  geom_boxplot()

# Apply Tukey's method to numeric columns
# amc_outliers <- boxplot.stats(df$amc, coef = 1.5)$out

# Identify which rows contain outliers
# amc_outlier_rows <- apply(df, 1, function(row) any(row %in% amc_outliers))

```

- we can see that the distribution of the amc is heavily right skewed
- we might need to do some transformation for better forecasting,
eg log transformation
- we have 6549 outliers and 14160 when the amc outliers are filtered out. 

#### Create a function for replacing outliers with median

```{r}
#create a function that replaces outliers with median values
replace_outliers_with_median <- function(x, coef = 1.5) {
  q1 <- quantile(x, 0.25)
  q3 <- quantile(x, 0.75)
  iqr <- q3 - q1
  fence_low <- q1 - coef * iqr
  fence_high <- q3 + coef * iqr
  x[x < fence_low | x > fence_high] <- median(x, na.rm = TRUE)
  return(x)
}
```


#### Imbalance in distribution of facilities

```{r}

df %>%
  group_by(group) %>%
  summarize(count = n_distinct(facility)) |> 
  gt()

```

- we have 12 facilities

#### Order Quantity Vs AMC

```{r}
df |> 
  ggplot(aes(log(order_quantity), log(amc))) +
  geom_point()
```


#### AMC per status


```{r}
df |> 
  ggplot(aes(status, amc)) + 
  geom_boxplot(fill = "lightblue") +
  # geom_jitter(aes(color = status), alpha = 0.1) +
  scale_y_log10() +
  labs(title = "Boxplots for AMC vs Product Status",
       y = "Average Monthly Consumption")

```

#### Order Quantity vs Status

```{r}
df |> 
  ggplot(aes(status, order_quantity)) + 
  geom_boxplot(fill = "lightblue") +
  # geom_jitter(aes(color = status), alpha = 0.1) +
  scale_y_log10() +
  labs(title = "Boxplots for Order Quantity  vs Product Status",
       y = "Order Quantity")
```


#### Most Ordered Products 

```{r}

df |> 
  group_by(product) |> 
  summarise(
    total_orders = sum(order_quantity)
  ) |> 
  arrange(desc(total_orders)) |> 
  head(6) |> 
  pull(product) -> top_ordered_products

df |> 
  group_by(product) |> 
  summarise(
    total_orders = sum(order_quantity)
  ) |> 
  arrange(desc(total_orders)) |> 
  head(6) |> 
  gt::gt()


```

#### total orders over time with outliers replaced by median

```{r}
df_subset <- df |> 
  filter(product %in% top_ordered_products)

df_subset$order_quantity <- replace_outliers_with_median(df_subset$order_quantity)



df_subset |> 
  group_by(product, year, month_range) |> 
  summarise(total_orders = sum(order_quantity), .groups = "drop") |> 
  ggplot(aes(month_range, total_orders, color = product)) +
  geom_line(aes(group = product)) +
  facet_wrap(~ year, ncol = 1) +
  # geom_text(aes(label = product)) +
  scale_y_continuous(labels = scales::number) +
  scale_color_discrete(name = "Product",
                       labels = function(x) str_sub(x, 1, 20)) +
  
  # theme(legend.position = "none")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

<!-- #### AMC along the reporting periods faceted by status -->

<!-- ```{r} -->

<!-- ggplot(df) + -->
<!--  aes(x = month_range, y = amc, fill = group, colour = group) + -->
<!--  geom_col() + -->
<!--  scale_fill_hue(direction = 1) + -->
<!--  scale_color_hue(direction = 1) + -->
<!--  theme_minimal() + -->
<!--  theme(legend.position = "right") + -->
<!--    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + -->
<!--  facet_wrap(vars(status),  -->
<!--  scales = "free_y", ncol = 1L) -->

<!-- ``` -->


- we have 1303 products
- some products do not span throughout the time frame, which is most likely due to the data quality issue



### Sample modelling
#### Subsetting for modelling

Now we focus on the business problem. that is to forecast the amc so as to calculate the amount needed to restock to 4 months.
 We are going to inspect the time series for the amc reported by group A on Paracetamol use

```{r}

df_selected <- df |> 
  filter(product == "Paracetamol Tablet 500 mg") |> 
  filter(group == "a")

# summarize the records for the selected group for reporting
df_selected <- df_selected |> 
  group_by(group, month_range, year) |>
  summarise(total_amc = sum(amc)) |> 
  arrange(year, month_range)

# replace outliers with median
df_selected$amc <- replace_outliers_with_median(df_selected$amc)

# creating time series object
ts_selected <- ts(df_selected$total_amc, frequency = 6, start = c(2020, 2.5))

autoplot(ts_selected)
ggAcf(ts_selected)
Box.test(ts_selected, type = "Lj")
```

The data seems to be mostly random white noise as the p-value and the Auto Correlation Function graph may suggest, so the amc as it is can not be used in for forecasting. Atleast for the data selected.
We might need to consider taking a different dataset to do the forecasting instead of the R&R reports, or after validating that we have all the data









#### Sample model comparison







#### Inspecting an Arima model fitted to the data selected

```{r}
fit <- auto.arima(ts_selected)
checkresiduals(fit)
summary(fit)
```

#### arima model on the data

```{r}
# seperate the data into a training set and a test set so as to validate our forecasts
train <- subset(ts_selected, end = (length(ts_selected) - 3))
#train our model and check accuracy for the predicted 3 following reporting periods
train |>
  auto.arima() |>
  forecast(h = 3) |>
  accuracy(ts_selected)

train |>
  auto.arima() |>
  forecast(h = 10) |>
  autoplot()
```

#### ETS model on the data

```{r}
train |>
  ets() |>
  forecast(h = 3) |>
  accuracy(ts_selected)

train |>
  ets() |>
  forecast(h = 10) |>
  autoplot()
```


#### Arima model with (2,1,2)

```{r}
train |>
  Arima(order = c(2,1,2), include.constant = T) |>
  forecast(h = 3) |>
  accuracy(ts_selected)

train |>
  Arima(order = c(2,1,2), include.constant = T) |>
  forecast(h = 10) |>
  autoplot()
```

ps: Arima(2,1,2) means we are using 2 autoregressive terms, 1 differentiation applied and 2 moving average terms in the model

##### using box cox transformation

```{r}
BoxCox.lambda(ts_selected)

train |>
  ets(lambda = 1) |>
  forecast(h = 3) |>
  accuracy(ts_selected)
```

lambda = 1 makes no difference as the time series is white noise

#### ets using log transformation

```{r}
train |>
  #log transform the data using lambda = 0
  ets(lambda = 0) |>
  forecast(h = 3) |>
  accuracy(ts_selected)

train |>
  ets(lambda = 0) |>
  forecast(h = 10) |>
  autoplot()

```

#### arima using log transformation

```{r}

train |>
  #log transform the data using lambda = 0
  auto.arima(lambda = 0) |>
  forecast(h = 3) |>
  accuracy(ts_selected)

train |>
  auto.arima(lambda = 0) |>
  forecast(h = 10) |>
  autoplot()
```


 - Using the new data we can see that the column names are also not consistent so this method would made more sense with a consistent API
- we do not have facility type when mbonea downloads the data









